Deterministic Bayesian inference methods for the Naomi model

### 2023-04-17 src/docs_18-04-23-lab-group/18-04-23-lab-group.Rmd
Finish slides for tomorrow #25

1. It runs quickly
2. It is flexible enough to be compatible with the model
3. We don't have better viable options

@@

* Ideally we want exact Bayesian inference: compute the posterior distribution of the parameters of the model given the data
* Computationally this amounts to solving a difficult integral
* We can't solve this, but we can give approximate answers

@@

* If we don't want to pretend the whole posterior distribution is Gaussian, another option is to pretend some of its marginals are
* This is how `TMB` works: it's up to the user to choose which parameters should be Gaussian using the `random` option

@@

* Gauss-Hermite quadrature is a method for picking nodes and weights based on the theory of polynomial interpolation
* It works especially well for statistical problems where the integrand looks like something multiplied by a Gaussian distribution
* The "adaptive" part means the nodes and weights are changed depending on the integrand -- this makes sense, especially when the integrand is also a function of the data
* Implemented by the `aghq` package [@stringer2021implementing]

@@

```{r, echo=FALSE, fig.cap="I attended one of the best engineering schools in Canada", out.width = '90%'}
knitr::include_graphics("the-plan.png")
```

@@

* For Malawi, Naomi has 24 hyperparameters: too many for a dense grid
* One proposed solution is to use principal components analysis (PCA) and keep only the first $s < 24$ dimensions
* We will use $k = 3$ points per dimension ($k = 2$ doesn't include the mode)

@@

```{r, echo=FALSE, message=FALSE, fig.cap="Illustration of what PCA-AGHQ looks like for the 2D example, keeping only the first principal component, $s = 1$."}
gg4 <- gg
mvQuad::rescale(gg4, m = mu, C = cov, dec.type = 1, levels = c(3, 1))

@@

* We run HMC for as long as it takes to get a good answer (around 3 days)
* Compare output of `TMB` with `aghq`: which is more similar to HMC?
* There are many ways you might compare posterior distributions
* We'll show Kolmogorov-Smirnov test results here

@@

Worse case TMB plot demonstrating KS test

@@

Overall KS results plot

@@

* Developed an approximate Bayesian inference method more accurate than `TMB` and faster than MCMC
* Scaled hyperparameter grid to >20 dimesions using PCA
* Relatively straightforward implementation using the `aghq` R package

@@

* More comprehensive inference comparison, particularly of model outputs
  * Maximum mean discrepancy
  * Pareto-smoothed importance sampling
* Extension to Laplace marginals via @wood2020simplified
  * May be more accurate, but have to think about computational costs

@@

* Working on a paper based on this work called "Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model" joint with Alex Stringer (Waterloo), Seth Flaxman (Oxford), Jeff Eaton (Imperial)
* Code and notebooks for this project are at `athowes.github.io/elgm-inf`

>>>>

1. It runs \textcolor{hilit}{quickly}
2. It is \textcolor{hilit}{flexible} enough to be compatible with the model
3. We don't have better viable options

@@

* Ideally we want exact Bayesian inference: compute the posterior distribution of the parameters of the model given the data
* Computationally this amounts to solving a difficult integral
* We can't solve this, but \textcolor{hilit}{we can give approximate answers}

@@

* If we don't want to pretend the whole posterior distribution is Gaussian, another option is to pretend \textcolor{hilit}{some of its marginals are}
* This is how `TMB` works: it's up to the user to choose which parameters should be Gaussian using the `random` option

@@

* \textcolor{hilit}{Gauss-Hermite quadrature} is a method for picking nodes and weights based on the theory of polynomial interpolation
* It works especially well for statistical problems where the integrand looks like something multiplied by a Gaussian distribution
* The \textcolor{hilit}{adaptive} part means the nodes and weights are changed depending on the integrand -- this makes sense, especially when the integrand is also a function of the data
* Implemented by the `aghq` package [@stringer2021implementing]

@@

```{r, echo=FALSE, fig.cap="One of Canada's top engineering schools (though ungraded)", out.width = '90%'}
knitr::include_graphics("the-plan.png")
```

@@

* For Malawi, Naomi has 24 hyperparameters: too many for a dense grid
* One proposed solution is to use \textcolor{hilit}{principal components analysis} (PCA) and keep only the first $s < 24$ dimensions
* We will use $k = 3$ points per dimension ($k = 2$ doesn't include the mode)

@@

```{r, echo=FALSE, message=FALSE, fig.cap="Illustration of what PCA-AGHQ looks like for the 2D example, keeping only the first principal component, $s = 1$."}
gg4 <- mvQuad::createNIGrid(2, "GHe", level = c(3, 1))
mvQuad::rescale(gg4, m = mu, C = cov, dec.type = 1)

@@

* We run HMC for as long as it takes to get a good answer (3 days)
* Compare output of `TMB` with `aghq`: which is more similar to HMC?

@@

```{r, echo=FALSE, fig.cap="Histogram of sample draws from each method for one latent field parameter. Which of \\texttt{TMB} and \\texttt{aghq} is closer to \\texttt{tmbstan}? Hard to say.", out.width = '75%'}
knitr::include_graphics("beta-alpha-histogram.png")
```

@@

```{r, echo=FALSE, fig.cap="The empirical cumulative distribution is slightly more enlightening.", out.width = '80%'}
knitr::include_graphics("ecdf-plot.png")
```

```{r, echo=FALSE, fig.cap="The Kolmogorov-Smirnov test statistic is the maximum difference between ECDFs.", out.width = '80%'}
knitr::include_graphics("ecdf-diff-plot.png")
```

```{r, echo=FALSE, fig.cap="Generating the KS test statistic for all latent field parameters, on average \\texttt{aghq} does better.", out.height = '80%'}
knitr::include_graphics("ks-density-plot.png")
```

@@

* Developed an approximate Bayesian inference method more accurate than `TMB` and faster than MCMC -- we're on the Pareto frontier!
* Scaled hyperparameter grid to >20 dimesions using PCA
* Relatively straightforward implementation using `aghq` R package, compatible with any model that has a `TMB` template

@@

* More comprehensive inference comparison, particularly of model outputs
  * Maximum mean discrepancy
  * Pareto-smoothed importance sampling
  * Simulation-based calibration
* Extension to Laplace marginals via @wood2020simplified
  * May be more accurate, but have to think about computational costs

@@

* Working on a paper based on this work called "Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model" joint with Alex Stringer (Waterloo), Seth Flaxman (Oxford), Jeff Eaton (Imperial)
* Code and notebooks for this project are available at `athowes.github.io/elgm-inf`
